A performance assessment of evolutionary algorithms in volunteer computing environments: the importance of entropy

### 2016-11-15 EvoPar2017/paper.Rnw
R4 Typos
Typos:

Section 4
Paragraph 1. 'Since the point of using volunteer' -> 'Since the point
of using a volunteer'
Paragraph 2. 'to the rest, specially' -> 'to the rest, especially'
Paragraph 5. 'The first thin we will' -> ''The first thing we will'
Paragraph 6. ''which so needs less' -> 'which needs less'

Section 5
Paragraph 1. 'This configuration achieve a' -> 'This configuration
achieved a'

graph2 & 1173561.5 & 6 & 4 & 761.5 & 22.0 & 0.0520509 & 36 & 73 & 129.04545\\
\hline
reboot & 796214.0 & 9 & 6 & 646.0 & 67.0 & 0.0955121 & 45 & 168 & 75.88889\\
\hline
cachereboot & 392917.0 & 12 & 12 & 576.0 & 79.0 & 0.1575092 & 55 & 225 & 43.28571\\
\hline
\end{tabular}
%% \end{tabular}
%% Initially generated with this code, 
% <<experiments, cache=FALSE,echo=FALSE>>=
%experiment.aggregates$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%kable(experiment.aggregates)
%@ %def 
\end{center}
\end{table}
%
%
%
Since the point of using volunteer computing system is to cut costs,
all experiments were set up using the OpenShift
PaaS, which provides a free tier, making the whole experiment cost
equal to nil. In fact, the NodIO system can be deployed to any
Platform or Infraestructure as a Service 
system as long as it can run node.js and can store logs on the same
place or in some other Log as a Service. Local storage can be used,
but it is not a requisite and, in fact, Heroku does not allow
persistent storage of files in its service. 
All these logs are available upon
request. Processed files are available in the same repository that
hosts this paper, as well as the scripts needed to process them.

%
%
\begin{figure}[htbp]
\centering
<<zipf, cache=FALSE,echo=FALSE,fig.height=6>>=
ggplot(ips.data,aes(x=idx,y=PUTs,group=Experiment,color=Experiment))+geom_point()+scale_y_log10()+ scale_colour_discrete(labels=c('base','graph1','graph2','reboot','cachereboot'))+scale_x_log10()
@ %def graph for IPs and puts
%
\caption{Log-log plot of ranked and normalized number of contributions (PUTs) per IP.\label{fig:zipf}}
\end{figure}
%
Every run had slightly different conditions, although in some cases it
was just a server reboot and a new round of announcements. Each run
batch included at least 30 completions, that is, evolution until the
solution was found. The number of runs was controlled by polling an
URL that indicated the number of runs made in the present batch so
far. Every volunteer contributed a certain number of individuals
(represented by the PUT HTTP request) to the server. The ranked number
of contributions per unique IP is shown in Figure \ref{fig:zipf} in a
log-log scale. Besides following the usual Zipf's law found in our
previous work and thus having a very similar appearance for all 5 sets, it
is interesting to see how some sets have a number of contributions per
IP that are slightly superior to the rest, specially so in {\tt
  graph2}. The last two sets, {\tt reboot} and {\tt cachereboot}, also
show a regime in the middle of the graph that shows that some users
are contributing more than in other similar sets. In general, these
differences in behavior are due more to the specific users that show
up in these experiments than to general laws, but it should also be
noted that those behaviors can be encouraged by design.

@@

%
\begin{table}
  \caption{Experiments with {\em overlapped} runs filtered out: summary of results. \label{tab:filtered}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Experiment & Median Time & \multicolumn{2}{|c|}{Median IPs} &\multicolumn{2}{|c|}{Median PUTs} & \#Runs &  Puts/IP \\
&   (ms)        & Total & Used & Total & Used &  & \\
\hline
base & 293685 & 13.0 & 8.0 & 483.0 & 57.0 & 47 & 35.50000\\
\hline
graph1 & 768053 & 12.0 & 5.0 & 601.0 & 79.0 & 21 & 45.84615\\
graph2 & 1477969 & 6.0 & 4.0 & 901.0 & 52.0 & 25 & 149.42857\\
\hline
reboot & 799471 & 9.5 & 6.5 & 656.5 & 67.0 & 44 & 76.48016\\
\hline
cachereboot & 463347 & 12.0 & 12.0 & 638.5 & 80.5 & 52 & 44.25911\\
\hline
\end{tabular}
%% Initially generated with this code, 
%%<<experiments.filtered, cache=FALSE,echo=FALSE>>=
%%experiment.aggregates.filtered$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%%kable(experiment.aggregates.filtered)
%%@ %def 
%
\end{center}
\end{table}
%
The first thin we will have to check is whether we have effectively eliminated these
overlaps. In order to do that, we will have to establish a threshold
under which we can reasonably expect that the solution cannot be found. In every
evolutionary algorithm it might happen, by chance, that the solution
is found in just a few evaluations, but that is usually not the
case. By looking at the logs in the last case we have established that
minimum as 16 PUTs; in less than 1600 generations no solution to that
difficult problem will be found. Let us filter then the runs,
eliminating those that used less than 1600 generations to find the
solution which can reasonably be said that they did so because some
browser ``carried over'' a chromosome from the previous run. The
results are shown in Table \ref{tab:filtered}. The first change is in
the column \#Runs, that shows that most of the runs in the initial
configurations found the solution thanks to these carried over
clients. However, in the last two rows, the ones that rebooted clients
when they detected changes in cache, the difference is minimal. This
means that one objective has been reached: if clients reboot when they
detect changes in the cache that imply a new run, overlaps are almost
eliminated. Going back to Table \ref{tab:runs}, we can see than the
ratio of effective contributions, those that feed the cache, increases
from around 13\% in the first case to almost 16\% in the last,
reaching lows of less than 6\% in some cases ({\tt graph2}). This
indicates that rebooting the clients has the positive effect of
increasing the contribution of clients to the common pool. However, we
should remark that the clients will still stop contributing if they
get stuck in a particular solution, so the ratio keeps being low
resulting in a slow change in the cache and a lower diversity of
clients that tap that cache. At any rate, an increased number of
effective PUTs will have a positive impact on the algorithm.

\begin{table}[htbp]
  \caption{Minute-wise correlation for number of participating volunteers. \label{tab:correlation}}
\begin{center}
<<correlations, cache=FALSE,echo=FALSE>>=
ip.correlations$Experiment <- c("base","graph1","reboot","cachereboot")
kable(ip.correlations)
@ %def correlations
\end{center}
\end{table}
%
We can check this impact by looking at the column that refers to
median time in Table \ref{tab:filtered} and its relation to the median
number of PUTs, which reflects the number of evaluations performed by
the clients. The last two rows have a low number of evaluations,
around 650, which is higher than the first two, probably due to the
lower volunteer supply, but better than the middle one; besides, the
amount is remarkably similar. But in the last case, it takes much less
time to find the solution {\em using the same number of
  evaluations}; the same situation applies if we compare {\tt graph1}
to {\tt cachereboot}. All variables are difficult to control in this
environment, but it might be due to the fact that parallelism has
increased, there are more volunteers contributing at the same time. In
fact, as it can be seen in the last column, the median number of
contributions per user is the lowest for {\tt cachereboot}, which so needs less contributions by users to
find the same solution. Besides, the last row shows that {\em all}
users actively contribute to the cache, while in the {\tt reboot}
strategy 30\% of users do not. This might be due to the fact that {\tt
  reboot} only does so when cache size is less than one, indicating
that this strategy is only partly successful in avoiding overlaps.

@@

Our main intention in this paper was to configure an evolutionary
algorithm in a volunteer computing environment to improve its
efficiency and eventually find out what this efficiency depends
on. The baseline configuration that we had used in other papers  included
{\em overlaps} or volunteers running islands that had actually started in
a previous experiment instance in the set. In previous papers, we were
interested in measuring the number of users or their behavior in bulk
and this overlap was not so important, but
in this case our focus was on the evolutionary algorithm so we had to
make all volunteers in it started more or less at the same time and
from a random population. In fact, we managed to do so, except in some
byzantine cases, by making the clients restart when the cache size
collapses to 1 or to a size much smaller than the current; we called
this client configuration {\tt reboot} and {\tt cachecrash}.  This
configuration achiever a positive secondary effect: all clients
contribute to fill the cache, having thus an actual effect on the
chromosome pool and effectively making all clients present contribute
to the simulation.

>>>>

graph2 & 1173561.5 & 6 & 4 & 761.5 & 22.0 & 0.0520509 & 36 & 73 & 129.04545\\
\hline
reboot & 796214.0 & 9 & 6 & 646.0 & 67.0 & 0.0955121 & 45 & 168 & 75.88889\\
\hline
cachereboot & 392917.0 & 12 & 12 & 576.0 & 79.0 & 0.1575092 & 55 & 225 & 43.28571\\
\hline
\end{tabular}
%% \end{tabular}
%% Initially generated with this code, 
% <<experiments, cache=FALSE,echo=FALSE>>=
%experiment.aggregates$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%kable(experiment.aggregates)
%@ %def 
\end{center}
\end{table}
%
%
%
Since the point of using a volunteer computing system is to cut costs,
all experiments were set up using the OpenShift
PaaS, which provides a free tier, making the whole experiment cost
equal to nil. In fact, the NodIO system can be deployed to any
Platform or Infraestructure as a Service 
system as long as it can run node.js and can store logs on the same
place or in some other Log as a Service. Local storage can be used,
but it is not a requisite and, in fact, Heroku does not allow
persistent storage of files in its service. 
All these logs are available upon
request. Processed files are available in the same repository that
hosts this paper, as well as the scripts needed to process them.

%
%
\begin{figure}[htbp]
\centering
<<zipf, cache=FALSE,echo=FALSE,fig.height=6>>=
ggplot(ips.data,aes(x=idx,y=PUTs,group=Experiment,color=Experiment))+geom_point()+scale_y_log10()+ scale_colour_discrete(labels=c('base','graph1','graph2','reboot','cachereboot'))+scale_x_log10()
@ %def graph for IPs and puts
%
\caption{Log-log plot of ranked and normalized number of contributions (PUTs) per IP.\label{fig:zipf}}
\end{figure}
%
Every run had slightly different conditions, although in some cases it
was just a server reboot and a new round of announcements. Each run
batch included at least 30 completions, that is, evolution until the
solution was found. The number of runs was controlled by polling an
URL that indicated the number of runs made in the present batch so
far. Every volunteer contributed a certain number of individuals
(represented by the PUT HTTP request) to the server. The ranked number
of contributions per unique IP is shown in Figure \ref{fig:zipf} in a
log-log scale. Besides following the usual Zipf's law found in our
previous work and thus having a very similar appearance for all 5 sets, it
is interesting to see how some sets have a number of contributions per
IP that are slightly superior to the rest, especially so in {\tt
  graph2}. The last two sets, {\tt reboot} and {\tt cachereboot}, also
show a regime in the middle of the graph that shows that some users
are contributing more than in other similar sets. In general, these
differences in behavior are due more to the specific users that show
up in these experiments than to general laws, but it should also be
noted that those behaviors can be encouraged by design.

@@

%
\begin{table}
  \caption{Experiments with {\em overlapped} runs filtered out: summary of results. \label{tab:filtered}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Experiment & Median Time & \multicolumn{2}{|c|}{Median IPs} &\multicolumn{2}{|c|}{Median PUTs} & \#Runs &  Puts/IP \\
&   (ms)        & Total & Used & Total & Used &  & \\
\hline
base & 293685 & 13.0 & 8.0 & 483.0 & 57.0 & 47 & 35.50000\\
\hline
graph1 & 768053 & 12.0 & 5.0 & 601.0 & 79.0 & 21 & 45.84615\\
graph2 & 1477969 & 6.0 & 4.0 & 901.0 & 52.0 & 25 & 149.42857\\
\hline
reboot & 799471 & 9.5 & 6.5 & 656.5 & 67.0 & 44 & 76.48016\\
\hline
cachereboot & 463347 & 12.0 & 12.0 & 638.5 & 80.5 & 52 & 44.25911\\
\hline
\end{tabular}
%% Initially generated with this code, 
%%<<experiments.filtered, cache=FALSE,echo=FALSE>>=
%%experiment.aggregates.filtered$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%%kable(experiment.aggregates.filtered)
%%@ %def 
%
\end{center}
\end{table}
%
The first thing we will have to check is whether we have effectively eliminated these
overlaps. In order to do that, we will have to establish a threshold
under which we can reasonably expect that the solution cannot be found. In every
evolutionary algorithm it might happen, by chance, that the solution
is found in just a few evaluations, but that is usually not the
case. By looking at the logs in the last case we have established that
minimum as 16 PUTs; in less than 1600 generations no solution to that
difficult problem will be found. Let us filter then the runs,
eliminating those that used less than 1600 generations to find the
solution which can reasonably be said that they did so because some
browser ``carried over'' a chromosome from the previous run. The
results are shown in Table \ref{tab:filtered}. The first change is in
the column \#Runs, that shows that most of the runs in the initial
configurations found the solution thanks to these carried over
clients. However, in the last two rows, the ones that rebooted clients
when they detected changes in cache, the difference is minimal. This
means that one objective has been reached: if clients reboot when they
detect changes in the cache that imply a new run, overlaps are almost
eliminated. Going back to Table \ref{tab:runs}, we can see than the
ratio of effective contributions, those that feed the cache, increases
from around 13\% in the first case to almost 16\% in the last,
reaching lows of less than 6\% in some cases ({\tt graph2}). This
indicates that rebooting the clients has the positive effect of
increasing the contribution of clients to the common pool. However, we
should remark that the clients will still stop contributing if they
get stuck in a particular solution, so the ratio keeps being low
resulting in a slow change in the cache and a lower diversity of
clients that tap that cache. At any rate, an increased number of
effective PUTs will have a positive impact on the algorithm.

\begin{table}[htbp]
  \caption{Minute-wise correlation for number of participating volunteers. \label{tab:correlation}}
\begin{center}
<<correlations, cache=FALSE,echo=FALSE>>=
ip.correlations$Experiment <- c("base","graph1","reboot","cachereboot")
kable(ip.correlations)
@ %def correlations
\end{center}
\end{table}
%
We can check this impact by looking at the column that refers to
median time in Table \ref{tab:filtered} and its relation to the median
number of PUTs, which reflects the number of evaluations performed by
the clients. The last two rows have a low number of evaluations,
around 650, which is higher than the first two, probably due to the
lower volunteer supply, but better than the middle one; besides, the
amount is remarkably similar. But in the last case, it takes much less
time to find the solution {\em using the same number of
  evaluations}; the same situation applies if we compare {\tt graph1}
to {\tt cachereboot}. All variables are difficult to control in this
environment, but it might be due to the fact that parallelism has
increased, there are more volunteers contributing at the same time. In
fact, as it can be seen in the last column, the median number of
contributions per user is the lowest for {\tt cachereboot}, which
needs less contributions by users to
find the same solution. Besides, the last row shows that {\em all}
users actively contribute to the cache, while in the {\tt reboot}
strategy 30\% of users do not. This might be due to the fact that {\tt
  reboot} only does so when cache size is less than one, indicating
that this strategy is only partly successful in avoiding overlaps.

@@

Our main intention in this paper was to configure an evolutionary
algorithm in a volunteer computing environment to improve its
efficiency and eventually find out what this efficiency depends
on. The baseline configuration that we had used in other papers  included
{\em overlaps} or volunteers running islands that had actually started in
a previous experiment instance in the set. In previous papers, we were
interested in measuring the number of users or their behavior in bulk
and this overlap was not so important, but
in this case our focus was on the evolutionary algorithm so we had to
make all volunteers in it started more or less at the same time and
from a random population. In fact, we managed to do so, except in some
byzantine cases, by making the clients restart when the cache size
collapses to 1 or to a size much smaller than the current; we called
this client configuration {\tt reboot} and {\tt cachecrash}.  This
configuration achieves a positive secondary effect: all clients
contribute to fill the cache, having thus an actual effect on the
chromosome pool and effectively making all clients present contribute
to the simulation.
